---
title: "BIEN Predicted Plant Species — Conservation Assessment"
output: html_document
params:
  species_df: NULL
  session_log: NULL
  polygon: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(sf)
```

<div class="alert alert-warning" style="border-left:6px solid #e67e22; background:#fff8e1; padding:16px 20px;">
<h4 style="margin-top:0;">&#9888; Mandatory Warnings — Read Before Use</h4>
<ol>
<li><strong>Modeled Predictions Only:</strong> All species predictions are derived from species distribution models (SDMs), not from field surveys. No species presence has been field-verified. Do not treat this list as a confirmed species inventory.</li>
<li><strong>Uneven Data Coverage:</strong> BIEN occurrence records are unevenly distributed across the Americas. Remote regions such as western Amazonia are substantially underrepresented in herbarium and field-survey databases. Absence from this list does not confirm absence from the landscape.</li>
<li><strong>Not for Legal or Policy Use:</strong> This tool must not serve as the sole basis for conservation assessments, environmental impact statements, land-use decisions, regulatory filings, or legal determinations. Independent expert review is required before operational use.</li>
<li><strong>Taxonomic Uncertainty:</strong> Species names follow the BIEN scrubbed taxonomy (aligned with The Plant List and related sources). These names may differ from other authoritative taxonomic treatments. Synonymy should be checked against a current reference before use.</li>
<li><strong>Native Status is Contextual:</strong> Native, introduced, and cultivated status determinations are context-dependent and not absolute. Do not rely on nativity assignments from this tool for policy purposes without independent expert review.</li>
</ol>
</div>

---

## Executive Summary

```{r exec-summary-data}
sp_df <- params$species_df
sl    <- params$session_log

total_sp    <- if (!is.null(sp_df)) nrow(sp_df) else 0L
tier_counts <- if (!is.null(sp_df) && nrow(sp_df) > 0) {
  table(sp_df$confidence_tier)
} else {
  table(character(0))
}

anchor_result  <- sl$anchor_check
n_anchor_pass  <- sum(anchor_result)
n_anchor_total <- length(anchor_result)

anchor_status <- if (n_anchor_pass == n_anchor_total) {
  "PASS"
} else if (n_anchor_pass == 0) {
  "FAIL"
} else {
  "PARTIAL"
}

anchor_color <- switch(anchor_status,
  PASS    = "green",
  FAIL    = "red",
  PARTIAL = "darkorange"
)
```

**Total predicted species:** `r total_sp`

**Anchor species plausibility check:**
<span style="background:`r anchor_color`; color:white; padding:3px 10px; border-radius:4px; font-weight:bold;">`r anchor_status`</span>
(`r n_anchor_pass` of `r n_anchor_total` anchor species present in predicted list)

**Species by confidence tier:**

```{r tier-table}
tier_order <- c("High", "Moderate", "Low", "Very Low")
tier_df <- data.frame(
  Tier  = tier_order,
  Count = vapply(tier_order, function(t) {
    if (t %in% names(tier_counts)) as.integer(tier_counts[[t]]) else 0L
  }, integer(1)),
  stringsAsFactors = FALSE
)
knitr::kable(tier_df, caption = "Predicted species count by confidence tier")
```

---

## Study Area

```{r study-area-data}
poly <- params$polygon
if (!is.null(poly)) {
  centroid  <- sf::st_coordinates(sf::st_centroid(sf::st_union(poly)))
  area_km2  <- sum(as.numeric(sf::st_area(poly))) / 1e6
  clon      <- round(centroid[1, "X"], 4)
  clat      <- round(centroid[1, "Y"], 4)
  area_disp <- round(area_km2, 1)
} else {
  clon      <- NA_real_
  clat      <- NA_real_
  area_disp <- NA_real_
}
```

| Field | Value |
|---|---|
| Centroid longitude | `r clon` |
| Centroid latitude | `r clat` |
| Area (km²) | `r area_disp` |
| Polygon source | `r sl$polygon_source` |
| Query timestamp (UTC) | `r sl$query_timestamp_utc` |
| Polygon SHA-256 | `r sl$polygon_sha256` |

---

## Methods

Species predictions are generated by intersecting the user-defined study area polygon with BIEN species range maps derived from MaxEnt species distribution models (SDMs). Range maps represent the predicted geographic extent of each species under a binary presence/absence threshold; the specific threshold applied by BIEN is range-dependent, and per-species AUC statistics are not available through the public API. Occurrence record counts are drawn from the BIEN occurrence database filtered to within the study area bounding box, restricted to records with `scrubbed_taxonomic_status == "Accepted"`, and excluding records with zero or missing coordinates (including likely data-entry artifacts at coordinate (0, 0)). Confidence tiers reflect the combined signal from range overlap percentage (proportion of the modeled range falling within the study area) and occurrence record count within the bounding box; tier thresholds are heuristic classifications and do not represent formal probability estimates or statistical confidence intervals. All results should be verified against independent data sources before use in any decision-making context.

---

## Occurrence Data Quality

The BIEN database aggregates occurrence records from herbarium collections, field surveys, plot networks, and other sources across the Americas. Sampling effort is highly heterogeneous: temperate North America and coastal or accessible regions are substantially better represented than remote tropical areas, including western Amazonia. Consequently, low occurrence counts in this report do not reliably indicate rarity — they may primarily reflect collection effort gaps rather than true ecological absence. Species with `data_support_n = 0` (zero occurrence records in the bounding box) are included solely on the basis of range map intersection and should be treated as highly speculative (`Very Low` confidence). The query used `natives.only = FALSE` to avoid artificially excluding range-edge or taxonomically under-curated native records; consequently, the species list may include records of uncertain native status.

---

## Species List

```{r species-list}
if (!is.null(sp_df) && nrow(sp_df) > 0) {
  display_df <- sp_df
  names(display_df) <- c("Scientific Name", "Family", "Confidence Tier",
                          "Occurrence Records", "Range Overlap (%)", "IUCN Status")
  knitr::kable(
    display_df,
    caption = paste0("Predicted plant species for study area (n = ", nrow(sp_df), ")"),
    digits  = 2
  )
} else {
  cat("No species were predicted for this study area.")
}
```

---

## Reproducibility Block

```{r repro-block}
if (!is.null(sl)) {
  scalar_fields <- sl[!names(sl) %in% "anchor_check"]
  scalar_df <- data.frame(
    Field = names(unlist(scalar_fields)),
    Value = as.character(unlist(scalar_fields)),
    stringsAsFactors = FALSE
  )

  anchor_df <- data.frame(
    Field = paste0("anchor_check.", names(sl$anchor_check)),
    Value = as.character(sl$anchor_check),
    stringsAsFactors = FALSE
  )

  repro_df <- rbind(scalar_df, anchor_df)
  knitr::kable(repro_df, col.names = c("Field", "Value"),
               caption = "Full session provenance log")
} else {
  cat("Session log not available (report rendered outside Shiny context).")
}
```

---

## References

- Maitner, B.S., Boyle, B., Casler, N., Condit, R., Donoghue, J., Durán, S.M., Guaderrama, D., Hinchliff, C.E., Jørgensen, P.M., Kraft, N.J.B., McGill, B., Merow, C., Morueta-Holme, N., Peet, R.K., Sandel, B., Schildhauer, M., Smith, S.A., Svenning, J.C., Thiers, B., Violle, C., Wiser, S., & Enquist, B.J. (2018). The BIEN R package: A tool to access the Botanical Information and Ecology Network (BIEN) database. *Methods in Ecology and Evolution*, 9(2), 373–379. DOI: 10.1111/2041-210X.12861 *(verify DOI before citing)*

- Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. *The R Journal*, 10(1), 439–446. DOI: 10.32614/RJ-2018-009

- `r paste(citation()$textVersion, collapse = " ")`
